Relational Parametricity for Algebraically Indexed Types

### 2012-06-01 paper/section-general-framework.tex
Nearly complete a full draft of the general framework section

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: an
\emph{index-erasure} semantics. The index-erasure semantics provides
FIXME. Relational semantics...

@@

We now define simultaneous substitutions of index terms. Given a pair
of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
\mathord: s_n$, a simultaneous substitution $\sigma : \Delta
\Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
such that for each $j$ such that $1 \leq j \leq n$, $\Delta \vdash e_j
: s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
\Delta \Rightarrow \Delta'$ and a variable $i_j \mathord: s_j$ in
$\Delta'$, we write $\sigma(i_j)$ for the index term $e_j$.

@@

We define a pair of simultaneous substitutions $\sigma, \sigma' :
\Delta \Rightarrow \Delta'$ to be judgmentally equal if their
component terms are judgmentally equal in the context $\Delta$: i.e.~
if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
$\Delta \vdash \sigma \equiv \sigma' : \Delta'$ when two simultaneous
substitutions are judgmentally equal.

@@

We now define the relational semantics of each well-indexed type
$\Delta \vdash A \isType$ as some binary relation on the index-erasure
interpretation of $A$. The relational interpretation that we define
will be a \emph{logical relation}, so the relational interpretation of
the function type will be the standard definition relating two
functions if they take related inputs to related outputs. We also use
the standard binary relational interpretations for the unit, tuple and
sum types. (We define the standard relational interpretations formally
below.)

It remains to devise the relational interpretation of primitive types
instantiated with index terms ($\tyPrim{X}{e_1,...,e_n}$), and
universally quantified types ($\forall i\mathord:s. A$). In the
index-erasure semantics, we assumed an assignment of a set
$\tyPrimSem{\tyPrimNm{X}}$ to each primitive type $\tyPrimNm{X} \in
\PrimTypeSet$. For the relational semantics of types
$\tyPrim{X}{e_1,...,e_n}$, we will assume some assignment of a binary
relation on $\tyPrimSem{\tyPrimNm{X}}$ to each primitive type
$\tyPrimNm{X}$ \emph{and} sequence of appropriately sorted (in some
index context $\Delta$) index terms $e_1,...,e_n$. Unlike the index
erasure interpretation, we do not assume a fixed relational
interpretation of the primitive types: in order to interpret the
universal quantification type,

FIXME: explain that we must deal with the fact that we wish to
quantify over extensions to the index context when we interpret the
universal quantification type.

For a set $X$, let $\Rel(X)$ be the set of binary relations $R
\subseteq X \times X$ on $X$.

Given an index context $\Delta$, a relation environment is a
(dependently typed) function:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{X})
\end{displaymath}
that respects equality on index terms: for any $\tyPrimNm{X} \in
\PrimTypeSet$ and pair of simultaneous substitutions $\sigma : \Delta
\Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma' : \Delta
\Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta \vdash
\sigma \equiv \sigma' : \primTyArity(\tyPrimNm{X})$, then $\rho\
\tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

\textbf{FIXME: define the composition of a relational environment with
  a simultaneous substitution}

\textbf{FIXME: reword this part} We define our relational semantics of
types with respect to a given collection of relation environments: we
assume that, for each index context $\Delta$, we are given a set of
relation environments $\relEnv{E}(\Delta)$. Note that this assignment
of sets of relation environments to index contexts is not assumed to
be compositional: it is not necessarily the case that the set of
relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is defined in
terms of $\relEnv{E}(\Delta_1)$ and $\relEnv{E}(\Delta_2)$. However,
it must satisfy the following two properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item \textbf{FIXME: update the formulation of this condition} If we
  have a pair of relation environments $\rho_1 \in \relEnv{E}(\Delta',
  i\mathord:s')$ and $\rho_2 \in \relEnv{E}(\Delta)$, along with an
  index context morphism $\sigma : \Delta' \Rightarrow \Delta$, such
  that the outer edges of the following diagram commute, for all
  primitive types $\tyPrimNm{X} \in \mathit{PrimType}$:
  \begin{displaymath}
    \xymatrix{
      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
      &
      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
      \\
      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
      &
      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
  bottom right of the diagram commute.
\end{enumerate}

\begin{example}[Units of Measure]
  FIXME: give two examples of sets of relational environments.
\end{example}

% FIXME: motivate this in terms of the semantics of forall types
Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be the following:
\begin{displaymath}
  \extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
\end{displaymath}

A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A \isType$ by
induction its derivation. Thus, we define a relation
$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
type.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now state and prove the abstraction theorem for well-typed programs
in our system. To state and prove this theorem for open terms, we
first need to extend the relational interpretation of types to typing
contexts. The relational interpretation of contexts is defined by the
following clauses, where we have used the pairing operation
$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & \isDefinedAs & \{(*,*)\} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \isDefinedAs & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}
The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpretating the application
of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

>>>>

We next present the semantics of types in
\autoref{sec:semantics-algebraically-indexed-types}. Types are first
interpreted without reference to the indexing terms: the
\emph{index-erasure} interpretation. The index-erasure interpretation
provides an ``underlying'' semantics to each type (and term), which we
think of as an abstraction of the actual operational behaviour of a
term. By erasing all the index terms, we can immediately see that the
run-time behaviour of programs cannot depend on index terms. We then
define the \emph{relational interpretation}, which interprets every
type as a binary relation on its index-erasure interpretation. The
relational interpretation acts as ``commentary'' on the underlying
interpretation, constraining the index-erasure interpretations that
are allowed. In particular, we use the relational interpretation to
enforce the invariance properties that we indentified in
\autoref{sec:motivating-examples}.

@@

It will technically convenient to express substitution of index terms
in our framework in terms of simultaneous substitutions.  Given a pair
of index contexts $\Delta$ and $\Delta' = i_1 \mathord: s_1, ..., i_n
\mathord: s_n$, a \emph{simultaneous substitution} $\sigma : \Delta
\Rightarrow \Delta'$ is a sequence of terms $\sigma = (e_1,...,e_n)$
such that for each $j$ such that $1 \leq j \leq n$, $\Delta \vdash e_j
: s_j$. Given a simultaneous substitution $\sigma = (e_1,...,e_n) :
\Delta \Rightarrow \Delta'$ and a variable $i_j \mathord: s_j$ in
$\Delta'$, we write $\sigma(i_j)$ for the index term $e_j$.

@@

We define a pair of simultaneous substitutions $\sigma, \sigma' :
\Delta \Rightarrow \Delta'$ to be judgmentally equal if their
component terms are judgmentally equal in the context $\Delta$: i.e.~
if $\Delta \vdash e_j \equiv e'_j : s_j$, for all $j$. We write
$\Delta \vdash \sigma \equiv \sigma' \Rightarrow \Delta'$ when two
simultaneous substitutions are judgmentally equal.

@@

We now define the relational semantics of each well-indexed type
$\Delta \vdash A \isType$ as some binary relation on the index-erasure
interpretation of $A$. For a set $X$, we write $\Rel(X)$ for the set
of binary relations $R \subseteq X \times X$ on $X$.

For the unit, tuple, sum and function types the we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types requires an interpretation of index
contexts. We assign interpretations to index contexts in terms of
\emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a
(dependently typed) function that, for each primitive type
$\tyPrimNm{X}$ and instantiation of its arguments $\sigma$, assigns a
binary relation $\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure
interpretation of $\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index term equality: for any
$\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous substitutions
$\sigma : \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ and $\sigma'
: \Delta \Rightarrow \primTyArity(\tyPrimNm{X})$ such that $\Delta
\vdash \sigma \equiv \sigma' \Rightarrow \primTyArity(\tyPrimNm{X})$, % FIXME: make sure the notation for ss equality is consistent
then $\rho\ \tyPrimNm{X}\ \sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

% FIXME: an example?

We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given a relational environment
$\rho \in \mathrm{RelEnv}(\Delta')$ and a simultaneous substitution
$\sigma : \Delta \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta)$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma \circ \sigma')$.

% FIXME: fix the notation here for interpretation of index contexts:
% should use some sort of semantic brackets

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.~ subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we can have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitution of index terms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an simultaneous substitution $\sigma
  : \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}$:
  %FIXME: check that these compositions are the right way round
  %FIXME: re-do this as a set of equations
  %FIXME: give a better English description of why this is needed
  \begin{displaymath}
    \xymatrix{
      {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
      \\
      {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
      &
      {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
      \\
      &
      &
      {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

Both of these conditions ensure that the relational interpretation of
types that we define below behaves correctly with respect to
simultaneous substitution of index terms
(\lemref{lem:tyeqsubst-relational}, below). The second condition may
seem mysterious at first, but it is essential for proving that the
relational interpretation of universal quantification types behaves
correctly with respect to application of simultaneous substitution.

\begin{example}[Units of Measure]
  We give two examples of assignments of sets of relational
  environments to index contexts in the units of measure example. The
  second example is a refinment of the first. FIXME: some general
  words about relational environments in these cases.
  \begin{enumerate}
  \item FIXME: assign a positive scaling factor to each variable in an
    index context, extend homomorphically to all index terms, and then
    use this to define a scaling relation
  \item FIXME: given an index context, assume a subgroup and a
    homomorphism into the scaling factors
  \end{enumerate}
\end{example}

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be
$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
= \rho \}$. The set of extensions of a relational environment will be
used in the interpretation of the universal quantification type.

A relational environment $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A \isType$ by
induction on its derivation: % Thus, we define a relation
% $\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-indexed
% type.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now state and prove the abstraction theorem for well-typed programs
in our system. To state and prove this theorem for open terms, we
first need to extend the relational interpretation of types to typing
contexts. The relational interpretation of contexts is defined by the
following clauses, where we have used the pairing operation
$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & \isDefinedAs & \{(*,*)\} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \isDefinedAs & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}
The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpreting the application
of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta \vdash' \Gamma \isCtxt$ and $\sigma : \Delta \Rightarrow
  \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
\end{lemma}
